Results

### 2016-02-08 doc/paper/results.Rmd
updated results text

There are two major components to the accuracy of a variant caller sensitivity and specificity. Sensitivity, describes how well a variant caller can identify variants that are known to be real and present in the viral population. A sensitivity of 1 indicates that all true positives are found. Specificity, on the other hand, measures how well a variant caller can correctly identify true negatives. For example, the influenza genome is slightly larger than 13KB. There are 3  potential variant nucleotides at each position leaving a variant caller with over 40,000 potential variants that can be called. The vast majority of these potential variants will not be present in the starting population. They will only be found as the result of a sequencing or RT-PCR error. A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1. A perfect variant caller will have a sensitivity and specificity of 1 meaning it can correctly identify all true variants and only identifies the true variants.

@@

There is a seemingly endless supply of variant callers on the market today, and it is beyond the scope of this work to give a detailed comparison of the accuracy of all these options. Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improved accuracy. In the process we hope to stress the importance of robustly benchmarking variant callers under the experimental conditions of a give study.

All variant callers attempt to distinguish true variants from errors by estimating accurate error rates across a given genome. These rates can take various metrics into account such as phred score, nucleotide position on a read, and the context of a given base call. The question then becomes is a putative variant found more often than would be expected given the applicable error rate. We have chosen DeepSNV as our variant caller of choice for this analysis. DeepSNV uses a clonal plasmid control to estimate an error rate for each possible nucleotide at each genomic position. The plasmid control contains is a clonal copy of the samples that are to be analyzed and is processed and sequenced in parallel. Thus the plasmid control allows DeepSNV to account for any context specific errors that may appear during the sequencing process. DeepSNV has been previously benchmarked using a cDNA mixture of plasmids containing a 1.5Kb regions of the HIV pol gene. This process involved minimal PCR and was devoid of a reverse transcription step. Patient derived RNA viral samples require an RT-PCR step, and the accuracy of this variant caller was previously unknown under these conditions. To characterize the accuracy of DeepSNV in the context of our sequencing pipeline, two lab strains, WSN33 and PR8, that differ at roughly 500 positions, were mixed at different frequencies to create known viral populations (`r miseq` A). Briefly, genomes were amplified separately in segment specific RT-PCR reactions. This was done to avoid biases in amplification that may skew frequency measurements towards one strain. Reconstituted cDNA genomes were made by mixing these products at equimolar ratios. The WSN33 mix was then serially diluted into PR8. These mixtures were sequenced using 2x250 reads on an Illumina Miseq and variants were called using DeepSNV.

>>>>

There are two major components that define the accuracy of a variant caller sensitivity and specificity. Sensitivity, describes how well a variant caller can identify variants that are known to be real and present in the viral population. A sensitivity of 1 indicates that all true positives are found. Specificity, on the other hand, measures how well a variant caller can correctly identify true negatives. For example, the influenza genome is slightly larger than 13Kb. There are 3  potential variant nucleotides at each position leaving a variant caller with over 40,000 potential variants that can be called. The vast majority of these potential variants will not be present in the starting population. They will only be found as the result of a sequencing or RT-PCR error. A variant caller that correctly identifies all potential false positives as true negatives has a specificity of 1. A perfect variant caller will have a sensitivity and specificity of 1 meaning it identifies all of the true variants an nothing else.

@@

There is a seemingly endless supply of variant callers on the market today, and it is beyond the scope of this work to give a detailed comparison of the accuracy of all these options. Instead we focus our efforts towards robustly benchmarking one variant caller and highlighting a few means for improved accuracy. In the process we hope to stress the importance of robustly benchmarking variant callers under the experimental conditions to which it is being applied.

All variant callers attempt to distinguish true variants from errors by estimating accurate error rates across a given genome. These rates can take various metrics into account such as phred score, nucleotide position on a read, and the context of a given base call. The question then becomes is a putative variant found more often than would be expected given the applicable error rate. We have chosen DeepSNV as our variant caller of choice for this analysis. DeepSNV uses a clonal plasmid control to estimate an error rate for each possible nucleotide at each genomic position. The plasmid control contains is a clonal copy of the genome that is to be analyzed and is processed and sequenced in parallel with the test samples. Thus the plasmid control allows DeepSNV to account for any context specific errors that may appear during the sequencing process. DeepSNV has previously been benchmarked using a cDNA mixture of plasmids containing a 1.5Kb regions of the HIV pol gene. This process involved minimal PCR and was devoid of a reverse transcription step. Patient derived RNA viral samples require an RT-PCR step, and the accuracy of this variant caller was previously unknown under these conditions. To characterize the accuracy of DeepSNV in the context of our sequencing pipeline, two lab strains, WSN33 and PR8, that differ at roughly 500 positions, were mixed at different frequencies to create known viral populations (`r miseq` A). Briefly, genomes were amplified separately in segment specific RT-PCR reactions. This was done to avoid biases in amplification that may skew frequency measurements towards one strain. Reconstituted cDNA genomes were made by mixing these products at equimolar ratios. The WSN33 mix was then serially diluted into PR8. These mixtures were sequenced using 2x250 reads on an Illumina Miseq and variants were called using DeepSNV.
